---
title: "Open Data - Discovering Opportunities"
author: "Sister Analyst"
date: "24/05/2023"
output: 
  html_document:
    theme: "yeti"
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = T,
                      include=TRUE,
                      prompt = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.height = 5,
                      fig.width = 7,
                      cache = FALSE)
```

## Introduction

We will create a data analysis report about air pollution in Nis with the explanation of the R code used. Data that will be used for the analysis comes from the stations, which are set up by an independent initiative [Vazduh Gradjanima](https://vazduhgradjanima.rs) 

Data is not frilly available with a single click, and we need to do some investigation. From the map <https://maps.sensor.community/#9/43.4655/21.3952> we can identify the klimerko stations in Nis and download their readings from <https://archive.sensor.community/2023-05-22/>. The readings are then combined into a single csv file 'data_nis.csv' that is made available from our GitHub. 

Once we have downloaded our data, we will start our workshop by installing [R](https://r-project.org) and [RStudio](https://www.rstudio.com), by following the instructions available on this link: [Install R i RStudia](https://instalirajr.netlify.app).

To learn how to use [R](https://cran.r-project.org/) and develop a report like this visit [Data Challenge Platform](https://datachallengewithr.rbind.io) with learning material developed in partnership with [UNDP Serbia](http://www.rs.undp.org). In particular we will focus on matirijal [Day 1](https://datachallengewithr.rbind.io/day1/) and [Day 2](https://datachallengewithr.rbind.io/day2/).

After we get familiar with the basic R syntax and data wrangling using data available from the open data portal Republic of Serbia https://data.gov.rs/sr/, we can dig into the air pollution data set following the instructions given below.  

## Air pollution case study

### Reading and organising data

We will start the analysis by uploading the necessary packages and data into R.

If you have not got the packages used in the following code, you will need to uncomment the first few lines (delete the `#` symbol) in the code below.

```{r}
#install.packages("rmarkdown")
#install.packages("leaflet")
#install.packages("lubridate")
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages("DT")

library(leaflet)
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyverse))

mydata <- read.csv('data/data_nis.csv', 
                   header=T) 
```

It is always a good idea to have a look at data you upload, before you start using it for your analysis.
```{r}
# scan data
glimpse(mydata)
```

We can do some tweaking, by removing a few variables and separating day from time and creating separate columns for latitude and longitude variables.
```{r}
# separate date from time
mydata <- separate(mydata, timestamp, c("date", "time"), sep = "T")
## remove the last character from the `time` variable
mydata$time <- (str_sub(mydata$time, end = -1))
# scan data
glimpse(mydata)
## remove 2nd and the last column
mydata <- mydata[, -c(2, 3)]
```

We will check how many unique records each variable in our data has.
```{r}
(uniq <- unlist(lapply(mydata, function(x) length(unique(x)))))
```

To us it is most interesting to notice that we have `10` stations in total. Next, we'll check if all of the stations are active, by examining the number of readings for each station.

```{r}
# how many reading each station makes
mydata %>%
  group_by(sensor_id) %>%
  summarise(no_readings = n()) %>%
  arrange(no_readings) %>% 
  DT::datatable()
```

There are all active and are providing the reading. 

### Mapping the stations

To see where the stations are allocated we will plot them on Google maps. We will use the `leaflet` package to do this and create a sub data with only a list of the stations and their positions.

```{r}
mydata[,1] <- as.factor(mydata[,1])
summary(mydata)
# creating a dataframe with only names and lat & lng for each station
stations <- data.frame(mydata[,c(1:3)]) %>% 
  drop_na()
stations [, 2] <- as.numeric(as.character(stations[,2]))
stations [, 3] <- as.numeric(as.character(stations[,3]))
summary(stations)
```

Once we have the subset data we can plot it using the `leaflet` package as below.

```{r}
minlat <- min(stations$lat)
maxlat <- max(stations$lat)
minlng <- min(stations$lon)
maxlng <- max(stations$lon)

stations %>% 
  group_by(sensor_id, lat, lon) %>% 
  leaflet() %>% 
  addTiles() %>%
  fitBounds(~minlng, ~minlat, ~maxlng, ~maxlat) %>% 
  addCircles(lng = ~lon, lat = ~lat,
             radius = 150, weight = 5, color = "black",
             fillColor = "red", fillOpacity = 0.7,  
             popup = ~paste("<b>", sensor_id)
  ) 
```

### Analysing the readings

In this section we will look through the readings and try to make some sense of it all.
If you are not familiar with the information collected by the stations, ie. about the pm particles you can check the following link:  <https://www.irceline.be/en/documentation/faq/what-is-pm10-and-pm2.5>.


#### Average concentration of particles each hour
```{r}
mydata$time <- factor(mydata$time)
mydata$time <- hms(as.character(mydata$time))

mydata %>% 
  group_by(hour(time)) %>% 
  summarise(no_readings = n()) %>% 
  DT::datatable()

mydata%>% 
  group_by(hour(time)) %>%
  summarise(mean_P1 = mean(P1, na.rm = TRUE), mean_P2 = mean(P2, na.rm = TRUE)) %>% 
  gather("pm_no", "mean_pm", -`hour(time)`, factor_key = TRUE) %>% 
  ggplot(aes(x = `hour(time)`, y = mean_pm, fill = pm_no )) +
  geom_bar(stat="identity", position = "dodge", color = "black") +
  coord_flip() +
  theme(plot.title = element_text(size = 14, vjust = 2, hjust=0.5)) +
  labs (title = "average value of P1 and P2 per hour", 
        caption = "Data from: https://vazduhgradjanima.rs", 
        x = "month", y = "average pm") +
  scale_fill_brewer(palette="Paired") + 
  theme(legend.position="bottom") 
```

